---
title: "aggregate"
author: "Ziyu Wu"
date: "2025-11-08"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. prepare packages
```{r}
dir.create("/home/azureuser/Rlibs", showWarnings = FALSE)
.libPaths("/home/azureuser/Rlibs")
.libPaths()
```

```{r}
library(ggplot2)
```

```{r}
library(patchwork)
```

```{r}
library(dplyr)
```

```{r}
library(tidyr)
```

```{r}
library(Rtsne)
```

```{r}
library(reshape2)
```

```{r}
library(viridis)
```

```{r}
library(scales)
```

```{r}
library(gridExtra)
```

```{r}
library(grid)
```

```{r}
library(rgl)
```

```{r}
library(smacof)
```

```{r}
library(purrr)
```

```{r}
library(remotes)
```

```{r}
library(keras3)
```

```{r}
library(viridis)
```

# 2. data

## 2.1 iris data

```{r}
# iris data
data(iris)
X_iris <-as.matrix(iris[, 1:4])
# X_iris <- unique(as.matrix(iris[, 1:4]))
X_iris_dist = as.matrix(dist(X_iris, method = "euclidean"))^2
# range(X_iris_dist[!diag(nrow(X_iris_dist))] )
# hist(X_iris_dist[!diag(nrow(X_iris_dist))] , breaks = 50)

```

```{r}
pca_res <- prcomp(X_iris, center = TRUE, scale. = TRUE)

pca_df <- data.frame(
  PC1 = pca_res$x[, 1],
  PC2 = pca_res$x[, 2],
  Species = iris$Species
)

var_explained <- summary(pca_res)$importance[2, 1:2] * 100

p <- (
  ggplot(pca_df, aes(x = PC1, y = PC2, color = Species, shape = Species)) +
    geom_point(size = 2.5, alpha = 0.9) +
    labs(
      x = paste0("PC1 (", round(var_explained[1], 1), "%)"),
      y = paste0("PC2 (", round(var_explained[2], 1), "%)")
    ) +
    theme_classic(base_size = 10) +
    theme(
      legend.position = "right",
      legend.title = element_blank()
    )
)

p

```


## 2.2 MNIST data
```{r}
mnist <- dataset_mnist()
```

```{r}
dim(mnist$train$x)
train_images <- mnist$train$x
train_labels <- mnist$train$y
```

# 3. functions
## 3.1 PMatrix & QMatrix function

```{r}
# get distance function
computeSquaredEuclideanDistance <- function(X,input = 'not distance') {
  if(input == 'not distance'){
  N <- nrow(X)
  DD <- matrix(0, N, N)
  for (i in 1:N) {
    for (j in i:N) {  
      d <- sum((X[i, ] - X[j, ])^2)
      DD[i, j] <- d
      DD[j, i] <- d
    }
  }}
  
  else DD = X^2
  return(DD)
}
# as.matrix(dist(X, method = "euclidean"))^2


# get beta(sigma) from perplexity for each point
distances2similarities <- function(D, n = -1, perplexity = 30, sigma = NULL) {
  
  N <- length(D)
  beta <- 1.0
  
  if (perplexity > 0) {
    # Binary search parameters
    min_beta <- -Inf
    max_beta <- Inf
    tol <- 1e-5
    max_iter <- 200
    iter <- 0
    found <- FALSE
    
    while(!found && iter < max_iter) {
      P <- exp(-beta * D)
      if (n >= 1) P[n] <- .Machine$double.eps  
      sum_P <- sum(P)
      H <- sum(beta * D * P) / sum_P + log(sum_P)
      
      Hdiff <- H - log(perplexity)
      
      if (abs(Hdiff) < tol) {
        found <- TRUE
      } else {
        if (Hdiff > 0) {
          min_beta <- beta
          if (is.infinite(max_beta)) {
            beta <- beta * 2
          } else {
            beta <- (beta + max_beta) / 2
          }
        } else {
          max_beta <- beta
          if (is.infinite(min_beta)) {
            beta <- beta / 2
          } else {
            beta <- (beta + min_beta) / 2
          }
        }
      }
      
      iter <- iter + 1
    }
  } else if (!is.null(sigma)) {
    beta <- 1 / (2 * sigma^2)
  }
  
  return(beta)
}


# get beta(sigma) vector
computeGaussianPerplexity <- function(X, perplexity = -1, sigma = 1.0, input = 'not distance') {
  
  N <- nrow(X)
  DD <- computeSquaredEuclideanDistance(X, input)
  beta <- numeric(N)
  
  if (perplexity < 0) {
    message("Using manually set kernel width")
  } 
  else {
    message("Using perplexity, not the manually set kernel width")
  }
  
  for (i in 1:N) {
    beta[i] = distances2similarities(DD[i, ], i, perplexity, sigma)
  }
   
  return(beta)
}


# get PMatrix
PMatrix <- function(X, perplexity, sigma, SN = 1, input = 'not distance') {
  N <- nrow(X)
  n = ncol(X)
  DD <- computeSquaredEuclideanDistance(X,input)
  beta <- computeGaussianPerplexity(X, perplexity, sigma,input)
  
  P <- matrix(0, nrow = N, ncol = N)
  
  # symmetric P matrix
  if(SN == 1){
    for (i in 1:N) {
    P[i, ] = exp(-beta[i] * DD[i, ])
    P[i, i] = 0
    P[i, ] = P[i, ] / sum(P[i, ])
  }
    P = (P + t(P)) / (2*N)
  } 
  
  # Pi|j
  else if(SN ==2){
    for (i in 1:N) {
    P[i, ] = exp(-beta[i] * DD[i, ])
    P[i, i] = 0
    P[i, ] = P[i, ] / sum(P[i, ])
  }
    P = P
  }
  
  return(P)
}

QMatrix <- function(Y,input='not distance') {
  N <- nrow(Y)
  DD <- DD <- computeSquaredEuclideanDistance(Y,input)
  Q <- 1 / (1 + DD)
  diag(Q) <- 0  
  Q <- Q / sum(Q)
  return(Q)
}


```

## 3.2 NE mds function

```{r}
NE_mds <- function(X, perplexity) {
  P <- PMatrix(X, perplexity,SN = 1) 
  # Z <- 1/max(P)
  Z = 1
  delta <- sqrt(1 / (P * Z) - 1)
  diag(delta) <- 0
  weight = P
  diag(weight) <- 0
  zinit <- cmdscale(as.matrix(dist(X)), k = 2 )
  #result <- fastmds(delta, 
  #                 w = weight, 
  #                 z = zinit,
  #                 rotate = FALSE, error.check = FALSE)
  result <- mds(delta, type = "ratio", weightmat = weight)
  return(result)
}
```

## 3.3 SKLAN mds function

```{r}
SKLAN_mds <- function(X, perplexity) {
  P <- PMatrix(X, perplexity, SN = 2) 
  N <- nrow(P)
  Delta <- matrix(0, nrow = N, ncol = N)
  P_log <- log(pmax(P, 1e-300))
  
  for (i in 1:(N-1)) {
    for (j in (i+1):N) {
      k_idx <- setdiff(seq_len(N), c(i, j))  
      diff_p <- P[i, k_idx] - P[j, k_idx]
      diff_log <- P_log[i, k_idx] - P_log[j, k_idx]
      
      delta_ij <- 0.5 * sum(diff_p * diff_log)
      
      Delta[i, j] <- delta_ij
      Delta[j, i] <- delta_ij
    }
  }
  
  #weight = P
  #diag(weight) = 0
  result <- mds(Delta, type = "ratio")
  #result <- fastmds(Delta,  
  #                  w = weight,  
  #                  rotate = FALSE, error.check = FALSE)
  return(result)
}
```

## 3.4 assessment function

```{r}
neighborhood_preservation <- function(a, b, k) {
  # a: high dimensional data distance
  # b: low dimensional data distance
  # k: neighborhood 
  n <- nrow(a)
  
  # rank matrix
  r <- matrix(0, n, n) 
  rhat <- matrix(0, n, n) 
  
  for (i in 1:n) {
    d <- a[i, ] 
    d[i] <- Inf 
    r[i, ] <- order(d) 
    } 
  
  for (i in 1:n) {
    d <- b[i, ] 
    d[i] <- Inf 
    rhat[i, ] <- order(d) 
    }
  
  # 1. Preservation
  lower <- n * k
  upper <- 0
  for (i in 1:n) {
    for (j in 1:k) {
      c <- rhat[i, j]
      if (c %in% r[i, 1:k]) {
        upper <- upper + 1
      }
    }
  }
  preservation <- upper / lower
  
  # 2.  Meta-criterion
  correct <- k / (n - 1)
  metacriterion <- preservation - correct
  
  # 3. Trustworthiness
  if (k < n/2) {
    GK <- n * k * (2 * n - 3 * k - 1)
  } else {
    GK <- n * (n - k) * (n - k - 1)
  }
  
  upper <- 0
  for (i in 1:n) {
    for (rank_low in 1:k) {
      j <- rhat[i, rank_low]
      rank_high <- which(r[i, ] == j)
      if (rank_high > k ) {
        upper <- upper + (rank_high - k)
      }
    }
  }
  
  trustworthiness <- 1 - 2 * upper / GK
  
  # 4. Continuity
  upper <- 0
  for (i in 1:n) {
    for (rank_high in 1:k) {
      j <- r[i, rank_high]
      rank_low <- which(rhat[i, ] == j)
      if (rank_low > k) {
        upper <- upper + (rank_low - k)
      }
    }
  }
  continuity <- 1 - 2 * upper / GK
  
  # 5. mean relative rank errors
  HK <- sum(abs(n - 2*(1:k)) / (1:k))
  
  sum_n <- 0
  sum_v <- 0
  
  for (i in 1:n) {
    for (rank_low in 1:k) {
      j <- rhat[i, rank_low]
      rank_high <- which(r[i, ] == j)
      sum_n <- sum_n + abs(rank_high - rank_low) / rank_high
      }
    }
  
  
  for (i in 1:n) {
    for (rank_high in 1:k) {
      j <- r[i, rank_high]
      rank_low <- which(rhat[i, ] == j)
      sum_v <- sum_v + abs(rank_high - rank_low) / rank_low
      }
    }
  
  MRRE_n <- 1- sum_n / (n * HK)
  MRRE_v <- 1- sum_v / (n * HK)
  
  
  
  
  # final results
  return(list(
    preservation = preservation,
    metacriterion = metacriterion,
    trustworthiness = trustworthiness,
    continuity = continuity,
    MRRE_n = MRRE_n,
    MRRE_v = MRRE_v
    
  ))
}
```

# 4 exploration

```{r}
t.sne.transformation.data <- function( perplexity = 200, z = 1.0 ,step = 0.025)
{ 
  x <- seq(0, 5, by = step)
  d = as.matrix(dist(x, method = "euclidean"))
  d_row = d[1,]
  d_row = d_row[-1]
  
  p = PMatrix(as.matrix(x), perplexity)
  p_row <- p[1, ] 
  p_row <- p_row[-1]
  
  q = QMatrix(as.matrix(x))
  q_row = q[1, ]
  q_row = q_row[-1]
  
  if(is.character(z) && z == "max"){ 
    z <- 1 / max(p_row)
  }
  
  z <- as.numeric(z)
  
  delta = sqrt(1/(p_row*z)-1)  
  
  return(list(d = d_row, p = p_row, q = q_row, delta = delta))
}
```

## 4.1 transformation in t-SNE
```{r}
perplexities <- c(10,30,50)
line_size <- 1
base_size <- 12

all_data <- lapply(perplexities, function(perp){
  data <- t.sne.transformation.data(perp, 'max')
  df <- data.frame(
    d = data$d,
    p_row = data$p,
    q_row = data$q,
    perplexity = perp
  )
  df_long <- df %>% 
    pivot_longer(cols = c(p_row, q_row), names_to = "type", values_to = "value")
  return(df_long)
}) %>% bind_rows()


all_data <- lapply(perplexities, function(perp) {
  data <- t.sne.transformation.data(perp, 'max')
  df <- data.frame(
    d = data$d,
    p_row = data$p,
    q_row = data$q,
    perplexity = perp
  )
  df_long <- df %>% 
    pivot_longer(cols = c(p_row, q_row), names_to = "type", values_to = "value")
  return(df_long)
}) %>% bind_rows()

perplexity_label <- function(variable, value) {
  paste0("perplexity = ", value)
}

p <- ggplot(all_data, aes(x = d, y = value, color = type)) +
  geom_line() +
  labs(x = "Euclidean Distance", y = "Probability Values") +
  scale_color_manual(values = c("blue", "red"), labels = c("pij", "qij")) +
  scale_x_continuous(limits = c(0, NA)) +
  scale_y_continuous(limits = c(0, NA), labels = scales::scientific) +
  theme_bw(base_size = base_size) +
  theme(
    axis.title = element_text(size = base_size + 2),
    axis.text = element_text(size = base_size),
    legend.title = element_blank(),
    legend.text = element_text(size = base_size),
    panel.grid.major = element_line(),
    panel.grid.minor = element_line()
  ) +
  facet_wrap(~perplexity, 
           ncol = 3,
           labeller = labeller(perplexity = function(x) paste0("perplexity = ", x)))
  

ggsave("tsne_transformation.png", plot = p, width = 10, height = 5, dpi = 600)

```

## 4.2 influence of Z in NE-based MDS
```{r}
base_size <- 12   
line_size <- 0.8  
z_values <- c(1, 100, "max")
perplexity_values <- c(10, 30, 50)


all_data <- expand.grid(
  z = z_values,
  perplexity = perplexity_values,
  stringsAsFactors = FALSE
) %>%
  pmap_dfr(function(z, perplexity) {
    
    tmp <- t.sne.transformation.data(
      perplexity = perplexity,
      z = z,
      step = 0.05
    )
    
    data.frame(
      d = tmp$d,
      delta = tmp$delta,
      z = as.factor(z),
      perplexity = as.factor(perplexity),
      panel = paste0("Perp=", perplexity, ", z=", z)
    )
  })


annot_data <- all_data %>%
  group_by(panel) %>%
  summarise(
    xmin = min(d),
    xmax = max(d),
    ymin = min(delta),
    ymax = max(delta),
    .groups = "drop"
  ) %>%
  mutate(
    x = xmin + 0.05 * (xmax - xmin),
    y = ymax - 0.05 * (ymax - ymin),
    label = paste0(
      "min = ", scientific(ymin, digits = 3),
      "\nmax = ", scientific(ymax, digits = 3)
    )
  )


p <- ggplot(all_data, aes(d, delta)) +
  geom_line(size = 0.9) +
  
  facet_wrap(
    ~ panel,
    ncol = 3,
    scales = "free"   
  ) +
  
  geom_label(
    data = annot_data,
    aes(x = x, y = y, label = label),
    hjust = 0,
    vjust = 1,
    size = 4.8,
    fill = "white",
    label.size = 0.3,
    inherit.aes = FALSE
  ) +
  
  labs(
    x = "High Dimensional Euclidean Distance",
    y = "NE-based MDS Delta"
  ) +
  
  theme_bw(base_size = 13) +
  theme(
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

p <- p +
  theme_bw(base_size = 16) +   
  theme(
    plot.title = element_text(size = 18, face = "bold"),
    strip.text = element_text(size = 16, face = "bold"),  
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 14),
    legend.title = element_text(size = 15),
    legend.text = element_text(size = 14)
  )

ggsave(
  "the influence of Z.png",
  p,
  width = 12,
  height = 10,
  dpi = 300
)
```



## 4.1 tsne descent

```{r}
x <- seq(0, 5, by = 0.1)
distx <- as.matrix(dist(x, method = "euclidean"))

y <- seq(0, 5, by = 0.1)
disty <- as.matrix(dist(y, method = "euclidean"))

p = PMatrix(as.matrix(x), perplexity = 30)
q = QMatrix(as.matrix(y))

p_row <- p[1, ]      
q_row <- q[1, ]      
distx_row <- distx[1, ]  
disty_row <- disty[1, ]  

n_x <- length(distx_row)
n_y <- length(disty_row)
result <- matrix(0, nrow = n_x, ncol = n_y)

for (i in 1:n_x) {
  for (j in 1:n_y) {
    result[i, j] <- 4* (p_row[i] - q_row[j]) * disty_row[j] / (1 + disty_row[j]^2)
  }
}

df <- expand.grid(
  x = distx_row[-1],
  y = disty_row[-1]
)
df$value <- as.vector(result[-1, -1])


p <- ggplot(df, aes(x, y, fill = value)) +
  geom_raster(interpolate = TRUE) +
  scale_fill_gradient2(low = "blue", mid = "green", high = "red", midpoint = 0) +
  labs(x = "High-dimensional Euclidean Distance Between i and j", 
       y = "Low-dimensional Euclidean Distance Between i and j") +
  theme_classic(base_size = 10) +
  coord_fixed()  


ggsave("tsne_gradient.png", plot = p, width = 6, height = 6, dpi = 300)


```


## 4.2 mds descent

```{r}
x <- seq(0, 5, by = 0.1)
distx <- as.matrix(dist(x, method = "euclidean"))

y <- seq(0, 5, by = 0.1)
disty <- as.matrix(dist(y, method = "euclidean"))

distx_row <- distx[1, ]  
disty_row <- disty[1, ]  

n_x <- length(distx_row)
n_y <- length(disty_row)
result <- matrix(0, nrow = n_x, ncol = n_y)

for (i in 1:n_x) {
  for (j in 1:n_y) {
    result[i, j] <- 2* (disty_row[j]-distx_row[i])
  }
}

df <- expand.grid(
  x = distx_row[-1],   
  y = disty_row[-1]    
)

df$value <- as.vector(result[-1,-1])  


p <- ggplot(df, aes(x, y, fill = value)) +
  geom_raster(interpolate = TRUE) +
  scale_fill_gradient2(low = "blue", mid = "green", high = "red", midpoint = 0) +
  labs(x = "High-dimensional Euclidean Distance Between i and j", 
       y = "Low-dimensional Euclidean Distance Between i and j") +
  theme_classic(base_size = 10) +
  coord_fixed()  

ggsave("MDS_gradient.png", plot = p, width = 6, height = 6, dpi = 300)
```

## 4.3 weighted mds descent
```{r}
x <- seq(0, 10, by = 0.1)
distx <- as.matrix(dist(x, method = "euclidean"))

y <- seq(0, 10, by = 0.1)
disty <- as.matrix(dist(y, method = "euclidean"))

p = PMatrix(as.matrix(x), perplexity = 30)
q = QMatrix(as.matrix(y))

p_row <- p[1, ]      
q_row <- q[1, ]      
distx_row <- distx[1, ]  
disty_row <- disty[1, ]  

n_x <- length(distx_row)
n_y <- length(disty_row)
result <- matrix(0, nrow = n_x, ncol = n_y)

for (i in 1:n_x) {
  for (j in 1:n_y) {
    result[i, j] <- 2* 
                    (dnorm(distx_row[i], mean = 0, sd = 5)) *
                    (disty_row[j]-distx_row[i])
  }
}

df <- expand.grid(
  x = distx_row[-1],   
  y = disty_row[-1]    
)

df$value <- as.vector(result[-1,-1])  


p <- ggplot(df, aes(x, y, fill = value)) +
  geom_raster(interpolate = TRUE) +
  scale_fill_gradient2(low = "blue", mid = "green", high = "red", midpoint = 0) +
  labs(x = "High-dimensional Euclidean Distance Between i and j", 
       y = "Low-dimensional Euclidean Distance Between i and j") +
  theme_classic(base_size = 10) +
  coord_fixed()  

ggsave("weighted_MDS_gradient.png", plot = p, width = 6, height = 6, dpi = 300)
```

# 5 plots

## 5.1 iris data

### 5.1.1 tsne(iris)

```{r}
set.seed(42)
perplexity_values <- 1:49
species_labels <- iris$Species

plot_list <- list()

for(p in perplexity_values){
  
  tsne_out <- Rtsne(
    X_iris,
    dims = 2,
    perplexity = p,
    max_iter = 500,
    verbose = FALSE,
    check_duplicates = FALSE
  )
  
  df_plot <- data.frame(
    X1 = tsne_out$Y[,1],
    X2 = tsne_out$Y[,2],
    Species = species_labels
  )
  
  g <- ggplot(df_plot, aes(X1, X2, color = Species)) +
  geom_point(size = 1) + 
  ggtitle(paste("Perplexity =", p)) +
  theme(
    plot.title = element_text(size = 8, face = "bold"),
    
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    
    legend.position = "right",    
    legend.title = element_text(size = 7),
    legend.text = element_text(size = 6)
  )
  
  plot_list[[p]] <- g
}

combined_plot <- wrap_plots(plot_list, ncol = 7, guides = "collect") &
  theme(legend.position = "right")

combined_plot

ggsave(
  filename = "tsne_perplexity_grid.png",
  plot = combined_plot,
  width = 14,
  height = 14,
  units = "in",
  dpi = 600
)
```

### 5.1.2 NE-based MDS(iris)

```{r}
set.seed(42)

perplexity_values <- 10:49
species_labels <- iris$Species

plot_list <- list()

i <- 1  

for(p in perplexity_values){
  
  NE_out <- NE_mds(X_iris, p)
  
  df_plot <- data.frame(
    X1 = NE_out$conf[,1],
    X2 = NE_out$conf[,2],
    Species = species_labels
  )
  
  g <- ggplot(df_plot, aes(X1, X2, color = Species)) +
    geom_point(size = 1) + 
    ggtitle(paste("Perplexity =", p)) +
    theme(
      plot.title = element_text(size = 8, face = "bold"),
      axis.title = element_blank(),
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      legend.position = "right",
      legend.title = element_text(size = 7),
      legend.text = element_text(size = 6)
    )
  
  plot_list[[i]] <- g
  i <- i + 1
}

combined_plot <- wrap_plots(plot_list, ncol = 5, guides = "collect") &
  theme(legend.position = "right")

combined_plot

ggsave(
  filename = "NE_perplexity_grid.png",
  plot = combined_plot,
  width = 14,
  height = 14,
  units = "in",
  dpi = 600
)
```

### 5.1.3 SKLAN-based MDS(iris)

```{r}
set.seed(42)

perplexity_values <- 10:49
species_labels <- iris$Species

plot_list <- list()

i <- 1   

for(p in perplexity_values){
  
  SKLAN_out <- SKLAN_mds(X_iris, p)
  
  df_plot <- data.frame(
    X1 = SKLAN_out$conf[,1],
    X2 = SKLAN_out$conf[,2],
    Species = species_labels
  )
  
  g <- ggplot(df_plot, aes(X1, X2, color = Species)) +
    geom_point(size = 1) + 
    ggtitle(paste("Perplexity =", p)) +
    theme(
      plot.title = element_text(size = 8, face = "bold"),
      axis.title = element_blank(),
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      legend.position = "right",
      legend.title = element_text(size = 7),
      legend.text = element_text(size = 6)
    )
  
  plot_list[[i]] <- g
  i <- i + 1
}

combined_plot <- wrap_plots(plot_list, ncol = 7, guides = "collect") &
  theme(legend.position = "right")

combined_plot

ggsave(
  filename = "SKLAN_perplexity_grid.png",
  plot = combined_plot,
  width = 14,
  height = 14,
  units = "in",
  dpi = 600
)
```

### 5.1.4 together(iris data)
```{r}
library(ggplot2)
library(patchwork)
library(Rtsne)

set.seed(42)
perplexity_values <- c(10,20,30,40)
species_labels <- iris$Species

plot_list <- list()

for (p in perplexity_values) {
  # t-SNE
  tsne_out <- Rtsne(X_iris, dims = 2, perplexity = p, verbose = FALSE, max_iter = 500,
                    check_duplicates = FALSE)
  df_tsne <- data.frame(
    Dim1 = tsne_out$Y[,1],
    Dim2 = tsne_out$Y[,2],
    Species = species_labels
  )
  p_tsne <- ggplot(df_tsne, aes(x = Dim1, 
                                y = Dim2, 
                                color = Species, 
                                shape = Species)) +
    geom_point(size = 2) +
    labs(title = paste("t-SNE, perplexity =", p), x = NULL, y = NULL) +
    theme_minimal(base_size = 12) +      
    theme(
      legend.position = "none",
      strip.text = element_text(size = 14),  
      axis.title = element_text(size = 14),
      axis.text = element_text(size = 14)
    )
  
  NE_out <- NE_mds(X_iris, perplexity = p)
  df_NE <- data.frame(
    Dim1 = NE_out$conf[,1], 
    Dim2 = NE_out$conf[,2],
    Species = species_labels
  )
  p_NE <- ggplot(df_NE, aes(x = Dim1, 
                                y = Dim2, 
                                color = Species, 
                                shape = Species)) +
    geom_point(size = 2) +
    labs(title = paste("NE-MDS, perplexity =", p), x = NULL, y = NULL) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "none",
          strip.text = element_text(size = 14),
          axis.title = element_text(size = 14),
          axis.text = element_text(size = 14))
  
  # SKLAN-MDS
  SKLAN_out <- SKLAN_mds(X_iris, perplexity = p)
  df_SKLAN <- data.frame(
    Dim1 = SKLAN_out$conf[,1],
    Dim2 = SKLAN_out$conf[,2],
    Species = species_labels
  )
  p_SKLAN <- ggplot(df_SKLAN, aes(x = Dim1, 
                                y = Dim2, 
                                color = Species, 
                                shape = Species)) +
    geom_point(size = 2) +
    labs(title = paste("SKLAN-MDS, perplexity =", p), x = NULL, y = NULL) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "none",
          strip.text = element_text(size = 14),
          axis.title = element_text(size = 14),
          axis.text = element_text(size = 14))
  
  plot_list <- c(plot_list, list(p_tsne, p_NE, p_SKLAN))
}

combined_plot <- wrap_plots(plotlist = plot_list, ncol = 3, guides = "collect") & 
  theme(legend.position = "bottom",
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 14))

ggsave("iris_comparison_plot.png", combined_plot, width = 12, height = 10, dpi = 300)


```

## 5.2 MNIST data

### 5.2.1 tsne(MNIST)

```{r}
# sample data
set.seed(42)
idx <- sample(1:dim(mnist$train$x)[1], 1000)
X_MNIST <- array_reshape(mnist$train$x[idx, , ],c(1000, 784))
X_MNIST <- X_MNIST / 255
X_MNIST_dist <- dist(X_MNIST, method = "euclidean")
label_MNIST <- mnist$train$y[idx]
```

```{r}
set.seed(42)

tsne_out_MNIST <- Rtsne(
  X_MNIST,
  dims = 2,
  perplexity = 30,
  verbose = TRUE,
  max_iter = 1000,
  check_duplicates = FALSE
)

```

```{r}
tsne_df <- data.frame(
  X = tsne_out_MNIST$Y[,1],
  Y = tsne_out_MNIST$Y[,2],
  label = factor(label_MNIST)
)

colors_alpha <- rainbow(10, alpha = 0.7)

p <- ggplot(tsne_df, aes(x = X, y = Y, color = label)) +
  geom_point(size = 0.5, alpha = 0.8) +
  scale_color_viridis_d(option = "turbo") + 
  labs(x = "t-SNE 1", y = "t-SNE 2", title = "MNIST t-SNE") +
  theme_minimal() 

print(p)


ggsave("MNIST_tSNE.png", plot = p, width = 6, height = 5, dpi = 300)


```

### 5.2.2 NE-based MDS(MNIST)

```{r}
set.seed(42)
NE_out_MNIST <- NE_mds(X_MNIST,perplexity = 30)
```

```{r}
set.seed(42)
ne_df <- data.frame(
  X = NE_out_MNIST$conf[,1],
  Y = NE_out_MNIST$conf[,2],
  label = factor(label_MNIST)
)

colors_alpha <- rainbow(10, alpha = 0.7)

p_ne <- ggplot(ne_df, aes(x = X, y = Y, color = label)) +
  geom_point(size = 0.5, alpha = 0.8) +
  scale_color_manual(values = colors_alpha) +
  labs(x = "NE-MDS 1", y = "NE-MDS 2", title = "MNIST NE-based MDS") +
  theme_minimal() 

print(p_ne)

ggsave("MNIST_NE_MDS.png", plot = p_ne, width = 6, height = 5, dpi = 300)

```

### 5.2.3 SKLAN(MNIST)
```{r}
set.seed(42)
SKLAN_out_MNIST <- SKLAN_mds(X_MNIST,perplexity = 30)
```

```{r}
sklan_df <- data.frame(
  X = SKLAN_out_MNIST$conf[,1],
  Y = SKLAN_out_MNIST$conf[,2],
  label = factor(label_MNIST)
)

colors_alpha <- rainbow(10, alpha = 0.7)

p_sk <- ggplot(sklan_df, aes(x = X, y = Y, color = label)) +
  geom_point(size = 0.5, alpha = 0.8) +
  scale_color_manual(values = colors_alpha) +
  labs(x = "SKLAN 1", y = "SKLAN 2", title = "MNIST SKLAN") +
  theme_minimal() 

print(p_sk)
ggsave("MNIST_SKLAN.png", plot = p_sk, width = 6, height = 5, dpi = 300)
```


```{r}
set.seed(42)
perplexity_values <- c(10,20,30,40)
colors_alpha <- rainbow(10, alpha = 0.7)

plot_list <- list()

for (p in perplexity_values) {
  tsne_out <- Rtsne(X_MNIST, dims = 2, perplexity = p, verbose = FALSE, max_iter = 500,
                    check_duplicates = FALSE)
  df_tsne <- data.frame(
    Dim1 = tsne_out$Y[,1],
    Dim2 = tsne_out$Y[,2],
    label = factor(label_MNIST)
  )
  p_tsne <- ggplot(df_tsne, aes(x = Dim1, 
                                y = Dim2, 
                                color = label)) +
    geom_point(size = 1, alpha = 0.8) +
    scale_color_viridis_d(option = "turbo") + 
    labs(title = paste("t-SNE, perplexity =", p), x = NULL, y = NULL) +
    theme_minimal(base_size = 12) +      
    theme(
      legend.position = "none",
      strip.text = element_text(size = 14),  
      axis.title = element_text(size = 14),
      axis.text = element_text(size = 14)
    )
  
  # NE-MDS
  NE_out <- NE_mds(X_MNIST, perplexity = p)
  df_NE <- data.frame(
    Dim1 = NE_out$conf[,1], 
    Dim2 = NE_out$conf[,2],
    label = factor(label_MNIST)
  )
  p_NE <- ggplot(df_NE, aes(x = Dim1, 
                                y = Dim2, 
                                color = label)) +
    geom_point(size = 1, alpha = 0.8) +
    scale_color_viridis_d(option = "turbo") + 
    labs(title = paste("NE-MDS, perplexity =", p), x = NULL, y = NULL) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "none",
          strip.text = element_text(size = 14),
          axis.title = element_text(size = 14),
          axis.text = element_text(size = 14))
  
  # SKLAN-MDS
  SKLAN_out <- SKLAN_mds(X_MNIST, perplexity = p)
  df_SKLAN <- data.frame(
    Dim1 = SKLAN_out$conf[,1],
    Dim2 = SKLAN_out$conf[,2],
     label = factor(label_MNIST)
  )
  p_SKLAN <- ggplot(df_SKLAN, aes(x = Dim1, 
                                y = Dim2, 
                                color = label)) +
    geom_point(size = 1, alpha = 0.8) +
    scale_color_viridis_d(option = "turbo") + 
    labs(title = paste("SKLAN-MDS, perplexity =", p), x = NULL, y = NULL) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "none",
          strip.text = element_text(size = 14),
          axis.title = element_text(size = 14),
          axis.text = element_text(size = 14))
  
  plot_list <- c(plot_list, list(p_tsne, p_NE, p_SKLAN))
}

combined_plot <- wrap_plots(plotlist = plot_list, ncol = 3, guides = "collect") & 
  theme(legend.position = "bottom",
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 14))

ggsave("MNIST_comparison_plot.png", combined_plot, width = 12, height = 10, dpi = 300)

```

# 6. assessment

## 6.1 three methods together (iris data)
```{r}
set.seed(42)

tsne_out <- Rtsne(X_iris, dims = 2, perplexity = 30, verbose = FALSE, max_iter = 500, check_duplicates = FALSE)
Y_dist_tsne <- as.matrix(dist(tsne_out$Y))^2
NE_out <- NE_mds(X_iris, perplexity = 40)
Y_dist_NE <- as.matrix(dist(NE_out$conf))^2
SKLAN_out <- SKLAN_mds(X_iris, perplexity = 40)
Y_dist_SKLAN <- as.matrix(dist(SKLAN_out$conf))^2

compute_metrics <- function(Y_dist, method_name) {
  k_value <- 2:100
  results <- data.frame(
    k = k_value,
    NP = NA,
    LCMC = NA,
    trustworthiness = NA,
    continuity = NA,
    MRRE_n = NA,
    MRRE_v = NA
  )
  
  for (k in k_value) {
    metrics <- neighborhood_preservation(X_iris_dist, Y_dist, k)
    results[results$k == k, 2:7] <- unlist(metrics)
  }
  
  results$Method <- method_name
  return(results)
}

tsne_metrics <- compute_metrics(Y_dist_tsne, "t-SNE")
ne_metrics <- compute_metrics(Y_dist_NE, "NE-based MDS")
sklan_metrics <- compute_metrics(Y_dist_SKLAN, "SKLAN-based MDS")

```


```{r}
metrics_order <- c("NP", "LCMC", "trustworthiness",
                   "continuity", "MRRE_n", "MRRE_v")

metric_labels <- c(
  NP = "Neighborhood Preservation",
  LCMC = "LCMC",
  trustworthiness = "Trustworthiness",
  continuity = "Continuity",
  MRRE_n = "MRRE_n",
  MRRE_v = "MRRE_v"
)

all_metrics <- rbind(tsne_metrics, ne_metrics, sklan_metrics)

all_metrics_long <- reshape2::melt(
  all_metrics,
  id.vars = c("k", "Method"),
  variable.name = "Metric",
  value.name = "Value"
)

all_metrics_long$Metric <- factor(all_metrics_long$Metric,
                                  levels = metrics_order)

plots <- lapply(metrics_order, function(m) {
  df <- all_metrics_long %>% filter(Metric == m)
  
  ggplot(df, aes(x = k, y = Value, color = Method)) +
    geom_line(linewidth = 1.5) +
    labs(
      x = "k",
      y = NULL,
      title = metric_labels[m]
    ) +
    scale_color_brewer(palette = "Set1") +
    theme_minimal(base_size = 20) +
    theme(
      plot.title = element_text(size = 24, hjust = 0.5),
      axis.title = element_text(size = 24),
      axis.text = element_text(size =24),
      legend.title = element_blank(),
      legend.text = element_text(size = 24)
    )
})

final_plot <- (plots[[1]] | plots[[2]] ) /
              (plots[[3]] | plots[[4]] )/
              (plots[[5]] | plots[[6]] ) +
              plot_layout(guides = "collect") &
              theme(legend.position = "bottom")


ggsave("iris_evaluation_metrics.png",
       plot = final_plot,
       width = 18,  
       height = 20, 
       dpi = 300)
```


## 6.2 three methods together (MNIST)
```{r}
set.seed(42)

tsne_out <- Rtsne(X_MNIST, dims = 2, perplexity = 30, verbose = FALSE, max_iter = 500, check_duplicates = FALSE)
Y_dist_tsne <- as.matrix(dist(tsne_out$Y))^2
NE_out <- NE_mds(X_MNIST, perplexity = 30)
Y_dist_NE <- as.matrix(dist(NE_out$conf))^2
SKLAN_out <- SKLAN_mds(X_MNIST, perplexity = 30)
Y_dist_SKLAN <- as.matrix(dist(SKLAN_out$conf))^2

compute_metrics <- function(Y_dist, method_name) {
  k_value <- c(1,100,200,300,400,500)
  results <- data.frame(
    k = k_value,
    NP = NA,
    LCMC = NA,
    trustworthiness = NA,
    continuity = NA,
    MRRE_n = NA,
    MRRE_v = NA
  )
  
  for (k in k_value) {
    metrics <- neighborhood_preservation(X_MNIST_dist, Y_dist, k)
    results[results$k == k, 2:7] <- unlist(metrics)
  }
  
  results$Method <- method_name
  return(results)
}

tsne_metrics <- compute_metrics(Y_dist_tsne, "t-SNE")
ne_metrics <- compute_metrics(Y_dist_NE, "NE-based MDS")
sklan_metrics <- compute_metrics(Y_dist_SKLAN, "SKLAN-based MDS")
```


```{r}
metrics_order <- c("NP", "LCMC", "trustworthiness",
                   "continuity", "MRRE_n", "MRRE_v")

metric_labels <- c(
  NP = "Neighborhood Preservation",
  LCMC = "LCMC",
  trustworthiness = "Trustworthiness",
  continuity = "Continuity",
  MRRE_n = "MRRE_n",
  MRRE_v = "MRRE_v"
)

all_metrics <- rbind(tsne_metrics, ne_metrics, sklan_metrics)

all_metrics_long <- reshape2::melt(
  all_metrics,
  id.vars = c("k", "Method"),
  variable.name = "Metric",
  value.name = "Value"
)

all_metrics_long$Metric <- factor(all_metrics_long$Metric,
                                  levels = metrics_order)

plots <- lapply(metrics_order, function(m) {
  df <- all_metrics_long %>% filter(Metric == m)
  
  ggplot(df, aes(x = k, y = Value, color = Method, linetype = Method)) +
    geom_line(linewidth = 1.5) +
    labs(
      x = "k",
      y = NULL,
      title = metric_labels[m]
    ) +
    scale_color_brewer(palette = "Set1") +
    theme_minimal(base_size = 20) +
    theme(
      plot.title = element_text(size = 24, hjust = 0.5),
      axis.title = element_text(size = 24),
      axis.text = element_text(size =24),
      legend.title = element_blank(),
      legend.text = element_text(size = 24)
    )
})

final_plot <- (plots[[1]] | plots[[2]] ) /
              (plots[[3]] | plots[[4]] )/
              (plots[[5]] | plots[[6]] ) +
              plot_layout(guides = "collect") &
              theme(legend.position = "bottom")


ggsave("MNIST_evaluation_metrics.png",
       plot = final_plot,
       width = 18,  
       height = 20, 
       dpi = 300)
```


